preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-02 Manuscript/manuscript.Rmd
edits to methods, results, and discussion

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution"
csl: styles.ref/genomebiology.csl
output:
  pdf_document: 
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associating Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by TAD calling algorithms. To aid in the precise identification of TAD boundaries we developed preciseTAD, a machine learning framework that leverages the spatial relationship of many high resolution ChIP-seq defined genomic elements, coupled with density-based clustering  and scalable partitioning techniques. Our framework precisely predicts chromosome-specific TAD boundaries on multiple cell types. We show that known molecular drivers of 3D chromatin including CTCF, RAD21, and SMC3 are more enriched at our predicted TAD boundaries compared to the boundaries identified by the popular ARROWHEAD TAD caller. Our results provide useful insights into the 3D organization of the human genome.

@@

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), less so for chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability. For MD, last part when ready to write an abstract]

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools ARROWHEAD [@durand2016juicebox] and Peakachu [@salameh2019supervised] respectively. ARROWHEAD TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines (Supplementary Table 1). The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site <!--(Supplementary Table 1)-->. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly<!--, Supplementary Table 2-->).

[???added spectralTAD]

@@

To create shifted binning, the first shifted bin was set to start at half of the resolution *r*, and continued in intervals of length *r* until the end of the chromosome (*mod r* + *r/2*). The shifted bins, referred hereafted as bins for simplicity, were then defined as boundary-containing regions (*Y = 1*) if they contained a TAD (or loop) boundary, and non-boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for classification (Figure 1).

@@

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was devised as a way for unbiased model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with one response vector, define column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

When evaluating performance across cell lines, a random forest model was built using ground truth boundaries and genomic annotations from one cell line such that the data matrix was constructed using genomic bins from $n-1$ chromosomes. The test data used to evaluate performance was constructed from the binned data of the $i^{th}$ holdout chromosome with genomic annotations from the other cell line. Thus, the models were trained on cell line specific data and tested on the other cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivies and specificities when tested on the cell line specific holdout chromosome test data, and reported the area under the curve (AUC).

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model was built on the optimal combination of data resolution (binning), predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_;!!!talk about in meeting; specify that each base pair has own prob; don't use "base pair" only "base" or "base level"]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, DBSCAN [??? We need to define what to italicize. Check below, make consistent throughout the manuscript;!!!talk about in meeting] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, preciseTAD then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

where $x_{ij}$ is the total number of base pairs of annotation $i$ in window $j$ and $w_{ij}$ is the number of base pairs that overlap between annotation $i$ and window $j$. The denominator, $L$, is the width of the windows (here, $L=50$). The rows of the heatmaps are ordered by enriched scores. The enriched scores are calculated as the sum of coverage values weighted by the distance to the flanked boundary, denoted by

@@

## Developing a machine learning framework for optimal TAD boundary prediction

We developed a machine learning framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and ($log_{2}$) Distance.](Figures/figure1.png)

We considered combinations of data from two cell line *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model.

@@

>>>>

---
title: "preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution"
csl: styles.ref/genomebiology.csl
output:
  pdf_document: 
  word_document:
    reference_docx: styles.doc/NIH_grant_style.docx
bibliography: References/references.bib
---

@@

Chromosome conformation capture combined with high-throughput sequencing experiments (Hi-C) have revealed that chromatin undergoes layers of compaction through DNA looping and folding, forming dynamic 3D structures. Among these are Topologically Associated Domains (TADs), which are known to play critical roles in cell dynamics like gene regulation and cell differentiation. Precise TAD mapping remains difficult, as it is strongly reliant on Hi-C data resolution. Obtaining genome-wide chromatin interactions at high-resolution is costly resulting in variability in true TAD boundary location by TAD calling algorithms. To aid in the precise identification of TAD boundaries we developed preciseTAD, a machine learning framework that leverages the spatial relationship of many high resolution ChIP-seq defined genomic elements, coupled with density-based clustering  and scalable partitioning techniques. Our framework precisely predicts chromosome-specific TAD boundaries on multiple cell types. We show that known molecular drivers of 3D chromatin including CTCF, RAD21, and SMC3 are more enriched at our predicted TAD boundaries compared to the boundaries identified by the popular ARROWHEAD TAD caller. Our results provide useful insights into the 3D organization of the human genome.

@@

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), less so for chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of bases [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning (ML) framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability. For MD, last part when ready to write an abstract]

@@

Within this study we used both TAD boundary and loop boundary data derived from the publicly available tools ARROWHEAD [@durand2016juicebox] and PEAKACHU [@salameh2019supervised] respectively. ARROWHEAD TAD boundaries were called from primary Hi-C data from Geo GSE63525 [@rao20143d] for the GM12878 and K562 cell lines. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for ARROWHEAD were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). PEAKACHU chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site. Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa]  (GRCh37/hg19 human genome assembly).

@@

To create shifted binning, the first shifted bin was set to start at half of the resolution *r*, and continued in intervals of length *r* until the end of the chromosome (*mod r* + *r/2*). The shifted bins, referred hereafted as bins for simplicity, were then defined as boundary-containing regions (*Y = 1*) if they contained a TAD (or loop) boundary, and non-boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for classification.

@@

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the SMOTE algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was devised as a way for unbiased model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

When evaluating performance across cell lines, a random forest model was built using ground truth boundaries and genomic annotations from one cell line such that the data matrix was constructed using genomic bins from $n-1$ chromosomes. The test data used to evaluate performance was constructed from the binned data of the $i^{th}$ holdout chromosome with genomic annotations from the other cell line. Thus, the models were trained on cell line specific data and tested on the other cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivies and specificities, and reported the area under the curve (AUC).

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model was built on the optimal combination of data resolution (binning), predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

We first construct a base level resolution predictor space for a particular chromosome **i**, using one of the above feature engineering procedures described above. We then validate our RF model on the base level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary point from the RF model, where $n_{i}$ represents the length of chromosome **i**. Next, DBSCAN (version 1.1-5) was applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close bases with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable bases with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, preciseTAD then implements an extension of partitioning around medoids (PAM), known as CLARA (**Clustering Large Applications**) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the **computeMatrix** and **plotHeatmap** tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

@@

where $x_{ij}$ is the total number of bases of annotation $i$ in window $j$ and $w_{ij}$ is the number of bases that overlap between annotation $i$ and window $j$. The denominator, $L$, is the width of the windows (here, $L=50$). The rows of the heatmaps are ordered by enriched scores. The enriched scores are calculated as the sum of coverage values weighted by the distance to the flanked boundary, denoted by

@@

## Developing a ML framework for optimal TAD boundary prediction

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base resolution.

We use called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publicly available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

In our ML framework boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png)

In total, we considered combinations of data from two cell lines *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model in an unbiased manner.

@@

**Supplementary Figure 9**. Density plots illustrating the genomic distance (in bases) between predicted TAD boundaries identified by preciseTAD for different combinations of thresholds (t) and eps values ($\epsilon$) on GM12878. The median genomic distance (range) is provided for each combination.
